Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-14 conclusion.tex
first draft of conclusion

In \cref{E-transmission}, I showed that this it is possible by appying two models.
The differing models, one statistical in nature and one mechanistic, each had their own strengths and weaknesses.
The statistical approach, based on the backcalculation framework, estimates incidence and makes fewer assumptions about the infection process.
However, the estimates are very uncertain.
The mechanistic approach, within the compartmental modelling framework, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.

These results highlighted the heterogeneity across regions and age groups in England.

These models relied on the theoretical work of \cref{E-inc-prev}.
This work developed the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
The major insight was the justification of the deterministic model framework and the importance of the distribution of the duration of positivity.

In \cref{E-perf-test,E-imperf-test}, I developed methodology to estimate the distribution of the duration of positivity from the CIS prevalence survey.
Therefore, showing that this important parameter can be estimated from prevalence surveys.
While I did make use of external estimates, from \cref{E-ATACCC}, this was to address weaknesses of the CIS study design that could be addressed in future studies (see \cref{conclusion:sec:future-work}).

Both approaches assumed that the distribution of the duration of positivity was perfectly known.

Estimating this distribution was the focus of the majority of this thesis.

@@

The feasibility of jointly modelling duration and incidence was not investigated in this thesis.
The model for duration in \cref{E-perf-test,E-imperf-test} estimated the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
Relaxing the uniformity assumption, would give incidence estimation.
The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterisation of the shape of the incidence curve would be major challenges in this extension.
Modelling frameworks could be adopted from the existing literature~\autocite[e.g.][]{taffeJoint,haySerosolver} used approximations (\eg empirical Bayes), smaller sample sizes, or high-performance computing to address these challenges.
However, the CIS is larger and incidence estimation desired at a more granular level than previously encountered.
Furthermore, high-performance computing cannot be used within the SRS.

@@

Other design questions are also posed by the work in this thesis.
For example, whether a representative sample of the population is optimal, or if certain groups should be oversampled.
Certain groups, such as younger individuals, were more likely to be infected.
It is plausible that sampling in proportion to incidence would be beneficial.
This is especially true if policy questions (\eg closing of schools) are of importance.
These questions could be investigated through simulation studies.

>>>>

In \cref{E-transmission}, I found that this it is possible, developing two models that can infer incidence and transmission from prevalence surveys.
One model was statistical in nature and the other mechanistic.
The statistical approach, based on the backcalculation framework, estimates incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
The mechanistic approach, within the compartmental modelling framework, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
These results highlighted the heterogeneity across regions and age groups in England.

The work in \cref{E-transmission} relied on the theoretical work of \cref{E-inc-prev}.
This work developed the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
The major insight was the justification of the deterministic model framework and the importance of the distribution of the duration of positivity.

These models both required knowledge of the distribution of the duration of positivity.
Estimating this distribution was the focus of the majority of this thesis.
In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated the CIS prevalence survey.
This did require external information from the ATACCC study (the results of \cref{E-ATACCC}), demonstrating that the CIS's design is not sufficient for estimating transmission.
However, future surveys may be able to avoid this issue (see \cref{conclusion:sec:future-work}).

Prior to this thesis, the duration of positivity in the general population was not well understood (see \cref{E-intro:sec:previous-duration-estimates}).
A particularly understudied area was quantifying the shape and size of the tail of the distribution, due to limited follow-up of individuals who test positive.

Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
A well-designed survey is sufficient to estimate all of the quantities of interest (as described in \cref{E-intro:sec:metrics}), except severity.

@@

The feasibility of jointly modelling duration and incidence was not investigated in this thesis.
The model for duration in \cref{E-perf-test,E-imperf-test} estimated the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
If, rather than assuming uniformity, the time of incidence was estimated, this would be a form of incidence estimation.
The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterisation of the shape of the incidence curve would be major challenges in this extension.
The basis of such a model could be existing modelling frameworks~\autocite[e.g.][]{taffeJoint,haySerosolver}.
These frameworks addressed the challenges by: taking approximations (\eg empirical Bayes), having smaller sample sizes, or using high-performance computing.
However, the CIS is larger and incidence estimation desired at a more granular level than previously encountered.
Furthermore, high-performance computing cannot be used within the SRS.

@@

Other design questions are also posed by the work in this thesis.
For example, whether a representative sample of the population is optimal, or if certain groups should be oversampled.
Certain groups, such as younger individuals, were more likely to be infected.
It is plausible that sampling in proportion to incidence would be beneficial.
This is especially true if policy questions (\eg closing of schools) are of importance.
These questions could be investigated through simulation studies.
In constructing such studies, it will be important to consider how these considerations might be different for different pathogens (\eg influenza, the most common source of respiratory pandemics) and tests (\eg rapid antigen tests, which have a shorter duration of positivity).

This thesis did not consider detection or early surveillance of emerging pandemics.
Initial uncertainty about local transmission in the UK may have led to suboptimal decisions at the start of 2020~\autocite{pellisChallenges}.
Community testing was not widely available at this stage of the pandemic, limiting the knowledge of the extent of local (as opposed to imported) transmission~\autocite{whittyTechReportCOVID}.
A representative survey of the population could have provided this information.
However, challenges in constructing a survey with the power to detect low levels of transmission, as well as the time required to set-up such a survey, would need to be addressed.

A final consideration is the estimation of the severity of a pathogen.
Severity of a pathogen is crucial for risk assessment, and can change over time~\autocite[e.g.][]{kirwanSeverityTrends}.
Two approaches could be taken to using a prevalence surveys for this purpose.
First, a cohort approach where the outcomes of individuals who test positive are followed up.
The low number of positive cases in the CIS would make this challenging.
In addition, the data would need to be linked to relevant records.
Second, a transmission model, such as the one in \cref{E-SEIR}, could jointly model severe events and prevalence.
Such an approach has previously been used~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2} but without the sophistication of the observation model developed in \cref{E-SEIR:sec:observation}.

### 2024-03-14 thesis.tex
Resolve to dos
fix subscripts for survival analysis
restructure sec 7.1
more detail in toc / navigation
first draft of conclusion

% Thesis-specific maths commands
\newcommand{\dmax}{d_\text{max}}
\newcommand{\psens}{p_\text{sens}}
\newcommand{\ntot}{n_\text{tot}}
\newcommand{\ndet}{n_\text{d}}
\newcommand{\nnodet}{n_\text{u}}
\newcommand{\pnodet}{p_\text{u}}
\newcommand{\Npop}{N_\text{pop}}
\newcommand{\Ncis}{N_\text{CIS}}
\newcommand{\ncis}{\vec{n_\text{CIS}}}
\newcommand{\na}{\vec{n_\text{a}}}
\newcommand{\pcis}{\vec{p_\text{CIS}}}
\newcommand{\sched}{\mathbb{T}}
\newcommand{\rate}{\mathfrak{R}}
\newcommand{\R}{\mathcal{R}_{0}}
\renewcommand{\Re}{\mathcal{R}_{e}(t)}
\newcommand{\Rc}{\mathcal{R}_{c}(t)}
\newcommand{\vlod}{v_\text{lod}}
\newcommand{\inform}{{_{\text{inform}}}}

@@

\setcounter{tocdepth}{1}

@@

@@

\chapter{Conclusion} \label{conclusion}
Household structure is not included in this model.
The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.

@@

\todo[inline]{Delete below lines - only here to make compiling individual chapters work}
\cref{perf-test:eq:piu}
\cref{inc-prev:eq:forward-substitute}
\listoftodos

>>>>

% Thesis-specific maths commands
\newcommand{\dmax}{d_\text{max}}
\newcommand{\psens}{p_\text{sens}}
\newcommand{\ntot}{n_\text{tot}}
\newcommand{\ndet}{n_\text{d}}
\newcommand{\nnodet}{n_\text{u}}
\newcommand{\pnodet}{p_\text{u}}
\newcommand{\Npop}{N_\text{pop}}
\newcommand{\Ncis}{N_\text{CIS}}
\newcommand{\ncis}{\vec{n_\text{CIS}}}
\newcommand{\na}{\vec{n_\text{a}}}
\newcommand{\pcis}{\vec{p_\text{CIS}}}
\newcommand{\sched}{\mathbb{T}}
\newcommand{\nsched}{n_{\sched}}
\newcommand{\rate}{\mathfrak{R}}
\newcommand{\R}{\mathcal{R}_{0}}
\renewcommand{\Re}{\mathcal{R}_{e}(t)}
\newcommand{\Rc}{\mathcal{R}_{c}(t)}
\newcommand{\vlod}{v_\text{lod}}
\newcommand{\inform}{{_{\text{inform}}}}

@@

\setcounter{tocdepth}{2}

@@

\todo[inline]{Mention use of early version of this work in ONS reports}

@@

\subfile{conclusion}

@@

\todo[inline]{Delete below lines - only here to make compiling individual chapters work}
\cref{perf-test:eq:piu}
\cref{inc-prev:eq:forward-substitute}
\cref{inc-prev:eq:EPt-matrix}
\cref{inc-prev:eq:Smatrix}
\listoftodos
